Constructing a universe for the setoid model

### 2020-06-22 paper.tex
More paper writing

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{proof}
\usepackage{tikz}
\usepackage{bbm}
\usepackage[bw]{agda}
\usetikzlibrary{cd}

\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}

@@

\author{
  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Ambrus Kaposi \inst{2}
\and
  Christian Sattler\inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Filippo Sestini\inst{1}
}

@@

...

Intensional Type Theory has nice computational properties, like decidability of
type-checking, that can make it more suitable for computer formalization, since
proof-checking can be implemented as an entirely automatic procedure.
%
However, a serious drawback of ITT is given by the lack of extensionality.
%
Two extensionality principles are particularly missed from intensional type
theory, in that they arguably constitute the bare minimum for type theory to be
comparable to set theory as a foundational system for set-level mathematics.
%
The first of these is functional extensionality, i.e. the principle according to
which two functions can be considered equal if they map equal inputs to equal
outputs.
%
The second principle is propositional extensionality, according to which two
propositions should be considered equal when logically equivalent.

These extensionality principles are not theorems of intensional type theory as
usually defined [cite], although they are consistent with it [cite].
%
As a consequence, these principles are often added as axioms when formalizing
mathematics using proof-assistants based on ITT. This solution, however, is far
from optimal, as the introduction of axioms destroys the nice computational
properties of the system, particularly canonicity. Unfortunately, one can show
that whenever equality is defined according to the usual inductive definition by
Martin-Loef, there is no hope to have a theory where both canonicity and
funext/propext coexist.

The first step out of this is to give up the idea of equality as a single
inductive definition. Instead, equality should be defined specifically for each
type former in the type theory. That is, every type former $A$ should be
introduced together with its own notion of equality $A^\sim$.
%
If one does this, then funext can be obtained by simply defining equality on
function types to be point-wise equality. That is, $(A \to B)^\sim\ f\ g :\equiv
\forall (x : A) , B^\sim\ (f x)\ (g x)$.

@@

This form of extensional reasoning is already available in type theory via
\emph{setoids}. A setoid is just a type $A$ (in this setting, we assume that all
types are sets in the HoTT sense) together with a propositional equivalence
relation $A^\sim : A \to A \to \Prop$.
%
One can thus reason extensionally in ITT by defining an internal notion of
setoid, and working with those instead of normal sets/types. Note that this
approach only works provided one ensures that every construction performed on
setoids respects the setoid relations involved. Unfortunately doing so is not an
easy task, and most constructions beyond the more trivial ones will generate a
great amount of proof obligations, especially when dependent setoids come into
play, leading to a situation commonly known as \emph{setoid hell}.

%% * Setoid TT: MLTT + setoid constructs

The good news is that all this work proving that our type theoretic
constructions respect the setoid relations is not necessary, because these
already respect setoid equality \emph{by construction}. More precisely, there
exists a sound interpretation of MLTT (with pi, sigma, etc...) into setoids,
that maps closed types/contexts to setoids and dependent types to indexed
setoids.
%
Thus there is no need to explicitly work with setoids and setoid relations in
TT, as one can simply carry out their constructions in plain MLTT and use the
aforementioned interpretation to map that construction into the world of setoid.

This is interesting in principle, but practically useless. Since our
constructions still take place in MLTT, we can't take advantage of the features
of the setoid model our syntactic constructions are mapped to, since they are
not accessible internall, from within the system itself.
%
This is where a key property of models comes into play. Models inform consistent
extensions of the type theory, based on what holds true in the model, by
reflecting the semanics into the syntax via additional syntactic constructs that
are then mapped back to the original notions via the semantics.

In our case, the existence of a setoid model suggests us ways to extend the
syntax of MLTT with syntactic constructs that correspond to the setoid
relations, allowing us to access and use them when proving theorems within the
theory.
%
This is the idea behind Setoid Type Theory, SeTT for short.

* Two incarnations of ITT that follow this approach are OTT and cubical TT. In OTT
... TODO. In cubical TT ... TODO. Note that we do not include HoTT in the
discussion since, although it is a version of ITT with extensional properties,
it's not yet known to enjoy canonicity.

... TODO ...

* universe of setoids

The current formulation of SeTT is already a satisfactory (\todo{maybe not the
  right word: satisfactory to who? according to what metric?}) presentation of
the setoid model, in that it includes all the features of the semantics that we
wished to reflect in the syntax. It is also arguably expressive enough to
formalize a good chunk of set-level mathematics. However, it is also currently
lacking in a serious way, as the lack of universes seriously limits its
introspection capabilities, and consequently its strength as a foundational
system. Note that SeTT already includes a (univalent) universe of strict
propositions; hence what we are referring to is a universe of setoids.

The inclusion of a universe of setoids in SeTT poses several interesting
questions. First of all, since a universe of setoids is itself a type in SeTT,
is must also be a setoid. However, it certainly cannot be a univalent one. One
question is therefore what kind of equality to equip such universe.
%
The second question is how one can justify such universe in the setoid model.
In particular what principles, if any, should be added to the metatheory in
order to be able to define the universe.

This paper is an attempt to give an answer to the questions above.
...TODO...

@@

By \emph{setoid model} we intend a class of models of (intensional) type theory
where contexts/closed types are interpreted as setoids, i.e. sets with an
equivalence relation, and dependent types are interpreted as dependent/indexed
setoids.
%
A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
to give a semantics to extensionality principles such as functional and
propositional extensionality. Here we consider a similar model construction, due
to T. Altenkirch \cite{setoid-model}.

The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory. The metatheory is also very simple, requiring only $\Pi$
types, $\Sigma$ types, and a universe of definitionally proof-irrelevant
propositions itself closed under such type formers.
%
Notably, the use of strict prop makes this model a strict one, namely one where
every equation in the model is a definitional equality in the metatheory.  The
model itself validates the extensionality principles mentioned above, like
funext, proext, and quotient types. As a consequence, this model construction
provides a way to explain extensionality, by giving a way to implement it from
stratch in terms of a base theory that is both very simple and intensional.

This idea of implementing extensionality via this model can be made both more
precise and more practical. Indeed, the strictness of this model implies that we
can turn this model into a syntactic translation, from which a syntactic
presentation of the model can be derived. (See next section, and [cite mpc'19])
\todo{rewrite this.}

\todo{mention that everything takes place in a type theoretic metatheory, with
  strict prop etc...}

@@

The setoid model has been recently adapted into Setoid Type Theory (SeTT), which
is justified by a syntactic translation into a very basic \emph{target theory}
\cite{mpc19}. This translation relies on the existence in the target theory of a
universe of definitionally proof-irrelevant propositions $\Prop$, as recently
implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
universe of propositions where equality of propositions is logical equivalence,
thus providing a very basic instance of univalence.

Technically, SeTT is an extension of MLTT with extra component to support setoid
reasoning. These extra components are essentially an internalization of the idea
that every type is a setoid. More precisely, we have rules [explaining TODO]
that contexts (aka closed types) are setoids, and that dependent types are
setoids indexed over setoids.

@@

%% Note that the setoid equivalence relations are strict propositions.

@@

Equality in SeTT differs from the standard approach by Martin-Loef in that every
type comes equipped with an equality relation. This is in contrast with ML's
identity type which is defined uniformly for any type.
%
Nevertheless, we can easily recover the usual ML identity type from setoid
equality:

\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]

Note that ... we have UIP... Moreover, contractibility of singletons holds by
UIP and by definition of equality on $\Sigma$ types. Therefore we can define the
$\J$ eliminator:

\[
\infer{
  \tmj{\GG}{\J_M\ p\ d}{M[a_0, a_1, p]}
}{
  \deduce{
    \tmj{\GG}{d}{M[a_0, a_0, \refl_A\ a_0]}
    \qquad
    \tmj{\GG}{p}{\underline{\Id_A\ a_0\ a_1}}
  }{
  \tyj{(\GG, x : A, y : A, p : \Id_A\ x\ y)}{M}
  }
}
\]

@@

\todo{Mention that the universe cannot be univalent}

\todo{Maybe mention the fact that it should be possible to give an elimination
  principle for the setoid universe. This also makes sense since it's not univalent.}

As currently defined, SeTT is a system already powerful enough to support the
formalization of part of set-level mathematics. Still, we would also like to
equip SeTT with a universe of setoids. As any type in SeTT, this universe must
in turn be a setoid. However, unlike the already existing universe of
propositions, a universe of setoids cannot be univalent, as this would force it
to be a groupoid. Instead, we must define its equivalence relation so that it
only reflects a simple, syntactic equality of codes.

As we use Type Theory as our metatheory, this universe of setoids is going to be
an internal construction in Type Theory. The usual way to encode universes in
Type Theory is by an inductive-recursive definition. As we will see in Section
TODO, it is indeed possible to give a IR definition of a universe of setoids.
%
However, we would prefer not to be forced to assume inductive-recursive types,
or IR in general, in the basic type theory which is the target of the setoid
translation. This is because the point of the translation is to explain a
complex system in terms of a very simple, well understood core, and we think
introducing IR would go against this goal.

Fortunately, it is possible to reduce some instances of induction-recursion to
plain induction (more specifically, inductive families), and the well known IR
universe construction is indeed one of them.

In the next section, we recall the IR construction of universes in type theory,
as well as lay out the process by which such construction can be turned to one
that only relies on inductive families. In Section TODO, we give a first IR
definition of the universe of setoids. After that we will see if and how such IR
definition can be turned into one that does not rely on IR.

@@

A standard, well-known way of encoding universes in type theory is via induction-recursion.
\todo{a couple of words about induction recursion}.

To define a universe, we simultaneously define the universe of codes itself, as
an inductive definition, and a family $\El : \U \to \Set$ that assigns, to each
code corresponding to some type, the meta-theoretic type of its elements.

For example, the definition of a small universe with booleans and $\Pi$ types
would be the following:

\begin{align*}
  & \textsf{data}\ \U\ : \Set_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

Note that $\U$ is defined to be in $\Set_1$ rather than $\Set$. This is
because... \todo{TODO}

As it turns out, induction-recursion is not really required to encode such
a universe, although is certainly makes the definition particularly nice.
%
In fact, we know that we can translate basic instances of induction-recursion
into inductive families using the equivalence of $I$-indexed families of sets
and sets over $I$, i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$.

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
any type that would have been obtained via $\El$ in the usual
inductive-recursive definition. As $\El$ is indexed on the type of codes, the
definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\Set_1$, rather than
$\Set$, since the definition of $\U$ quantifies over all possible sets. Thus,
the meta-theory must contain at least one universe for this construction to go
through.

@@

\todo{QIIT II LICS paper. this part is already in the HIIT paper}

We given an inductive definition of $\setoidU$, mutually with a recursive
definition of everything else.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{tikz}
\usepackage{bbm}
\usepackage[bw]{agda}
\usetikzlibrary{cd}

\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\mType}{\mathbf{Type}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\mProp}{\mathbf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Sub}{\mathrm{Sub}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}
\newcommand{\Uty}{\mathds{U}}
\newcommand{\EL}{\mathds{E}l}

@@

\author{
  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Simon Boulier
\and
  Ambrus Kaposi \inst{2}
\and
  Christian Sattler\inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Filippo Sestini\inst{1}
}

@@

A central aspect of Type Theory is the coexistence of two notions of
equality. On the one hand definitional equality, the computational equality that
is built into the formalism. On the other hand propositional equality, the
internal notion of equality that is actually used to state and prove equational
theorems within the system. The precise balance between these two notions is at
the center of type theory research; however, it is generally understood that to
properly support formalization of mathematics in type theory, one should aim for
a definitional equality that is decidable and therefore machine-checkable, and a
notion of propositional equality that is as extensional as possible.

Two extensionality principles seem particularly desirable, in that they arguably
constitute the bare minimum for type theory to be comparable to set theory as a
foundational system for set-level mathematics, in terms of power and ergonomics.
One is function extensionality, according to which functions are equal if
point-wise equal. Another is propositional extensionality, equating all
propositions that are logically equivalent.

Extensional Type Theory (ETT) does support extensional reasoning to some degree,
thus seemingly being a convenient setting for formalization of
mathematics. However, equality reflection makes type-checking ETT terms
computationally unfeasible, if not impossible.

On the other hand, Intensional Type Theory has nice computational properties
that can make it more suitable for computer implementation, but as normally
defined [cite ML paper?] it lacks the amount of extensionality necessary to
effectively \emph{use} it.
%
It is known from model constructions that extensional principles like function
extensionality (funext) are consistent with ITT.
%
Moreover, ITT extended with uniqueness of identity proofs (UIP) and funext is
known to be as powerful as ETT [cite hofmann]. These principles can be added as
axioms, however destroying some computational properties of ITT such as
canonicity.

Ultimately, what is needed is a formulation of ITT that supports the
aforementioned extensionality principles, while retaining convenient
computational properties like canonicity.
%
Unfortunately, one can show that whenever identity types are defined according
to Martin-Loef's inductive definition, there is no hope to have canonicity and
funext coexist.
%
The first step towards a solution is to give up the idea of equality as a single
inductive definition. Instead, equality should be defined specifically for each
type former in the type theory. According to this principle, funext can be
justified by simply defining equality on function types to be point-wise
equality.

@@

Sets, or types, together with a propositional equivalence relation as their
notion of equality are known as \emph{setoids}. There are several definitions of
the setoid model of Type Theory, the first of which was constructed to show
consistency of funext for ITT [cite]. This consistency result is implied by the
observation that UIP and funext hold in the setoid model. Moreover, it was shown
by Altenkirch [cite] that is the model construction is carried out in a
metatheory with a universe of strict (definitionally proof-irrelevant)
propositions, it is possible to define a univalent universe of propositions,
therefore satisfying propositional extensionality. The setoid model thus
satisfies all the extensionality principles that we would like to extend ITT
with. The question is whether there exists a version of ITT that supports setoid
reasoning, and hence the forms of extensionality enables by it.

This question was revisited and answered in Altenkirch et al. [cite]. In this
paper, the authors define SeTT an extension of intensional MLTT + Prop with
constructs for setoid reasoning, where funext and propext hold by
definition. Moreover, they justify SeTT via a syntactic translation, after
observing that this is made possible by the strictness of the model (every
equation holds definitionally). This is in contrast with other type theories
based on the setoid model, like OTT [cite] and XTT [cite], which instead rely on
ETT as metatheory for their justification. A major contribution of SeTT is
therefore to provide a way to bootstrap extensionality, by translating it into a
small core intensional theory by means that do not require extensionality to
begin with.

The current formulation of SeTT is already a rich theory. However, its
introspection capabilities are currently lacking, as they are limited to the
univalent universe of propositions. We would like to internalise the notion of
type in SeTT, thus extending the theory with a universe of setoids.
%
This brings up several questions, one of which has to do with the notion of
equality with which the universe should come equipped. The universe of setoids
is itself a setoid (as any type is), so it certainly cannot be a univalent
universe, since a univalent setoid would be a groupoid rather than a setoid.
%
Another question asks how one can justify such universe in the setoid model. In
particular what principles, are needed from the metatheory in order to define
the universe.

\paragraph{Contributions}

This paper documents our work towards the construction of a universe of setoids
inside the setoid model, and tries to answer these and other questions related
to the design and implementation of this construction. Our main contribution is
the definition of the universe in the setoid model; this is initially given as
an inductive-inductive-recursive definition, which is then translated firstly to
an inductive-inductive definition, and secondly to a purely inductive type. As a
consequence, we show that we only need to assume indexed W-types in the
metatheory (along with some obligatory basic tools like $\Sigma$ and $\Pi$
types) to justify the universe.

\paragraph{Structure of the paper}

In Section~\label{sec:metatheory}...

\section{$MLTT^{Prop}$}\label{metatheory}

This section describes $MLTT^{Prop}$, the metatheory in which we model
construction takes place. It is crucial to keep the metatheory as minimal as
possible.

$MLTT^{Prop}$ is an intensional MLTT with $\Pi, \Sigma, \textsf{Bool},
\mathbf{0}, \mathbf{1}$, and a universe of strict propositions $\mProp$. We also
have definitional $\eta$ rules for $\Pi, \Sigma$, and $\mathbf{1}$
types. Propositions can be lifted to types via $\textsf{Lift} : \mProp \to
\mType$. The universe of propositions is inhabited by false $\bot$ and
propositional truncation $\|\_\| : \mType \to \mProp$. We also require indexed
W-types, both in $\mType$ and $\mProp$:

\begin{align*}
  W & : (S : \mType) \to (S \to \mType) \to \mType \\
  W_p & : (S : \mType) \to (S \to \mType) \to \mProp
\end{align*}

We also require the propositional versions of $\mathbf{1}$, $\Pi$, and $\Sigma$,
however these can be defined from their $\mType$ counterparts via truncation:

\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

Note that a universe of strict propositions has been recently added to the Agda
proof assistant [cite], thus making $MLTT^{Prop}$ a subset of Agda. In fact,
most of the developments presented here have been formalized and proof-checked
using Agda extended with strict propositions. For convenience we will employ
Agda notation for metatheoretic constructions throughout the paper, with the
implication that these can be understood as, or easily translated to, equivalent
$MLTT^{Prop}$ definitions.

@@

By \emph{setoid model} we intend a class of models of (intensional) type theory
where contexts/closed types are interpreted as setoids, i.e. sets with an
equivalence relation, and dependent types are interpreted as dependent/indexed
setoids.
%
A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
to give a semantics to extensionality principles such as functional and
propositional extensionality. Here we consider a similar model construction, due
to T. Altenkirch \cite{setoid-model}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory, in our case $MLTT^{Prop}$.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates extensionality
principles like funext, proext, and quotient types. As a consequence, it
provides a way to bootstrap and explain extensionality by implementing it in a
core intensional type theory.

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers. The core structure can be defined as the following signature:

\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

Contexts are given by setoids, that is types together with an equivalence relation.

\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & ...
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
(strict) category of setoids. The definition can be unfolded as follows:

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \dots
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

This definition of types in the setoid model is different from the original one
in [cite LICS 99 paper], but it is equivalent to it [cite].
%
Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
fibration $\GG . A \to \GG$. Note that we only need to include components for
the functorial action on objects and morphisms, since the functor laws follow
from proof-irrelevance.

\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]

\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : ...
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]

As defined, we can show that the setoid model validates the usual basic type
formers ($\Pi, \Sigma$, etc.), including function extensionality and a universe
of strict propositions satisfying propositional extensionality [cite LICS
  99]. Note that we don't need neither identity types nor inductive types
(W-types) in the metatheory to achieve this. However, W-types are needed when
constructing the setoid universe in the next sections.

@@

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.

A setoid translation is used in \cite{mpc19} to justify Setoid Type Theory
(SeTT), an extension of $MLTT^{Prop}$ (without W-types) with equality types for
contexts and dependent types that reflect the setoid equality of the model in
the syntax.

We recall the rules of SeTT that extend $MLTT^{Prop}$ below, but with a
variation. Whereas the equality types in \cite{mpc19} are stated as elements of
the universe of propositions internal to SeTT itself, we here state the context
equalities as elements of the external, metatheoretic universe
$\mProp$. Equality on dependent types is defined as before.

We have a universe of propositions $\Prop$ defined as follows:

\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\]

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. As in the model, equality for dependent types is indexed over context
equality.

@@

We have rules witnessing that these are indeed equivalence relations. We only
recall reflexivity:

\[
\infer{
  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
}{
  \rho : \Sub\ \GD\ \GG
}
\qquad
\infer{
  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
}{
  A : \Ty\ \GG
  \qquad
  \rho : \Sub\ \GD\ \GG
  \qquad
  a : \Tm\ \GD\ A[\rho]
}
\]

@@

Notably, equality constructors in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%
\begin{align*}
  & (\Pi\ A\ B)^\sim\ p\ f_0\ f_1 = ... \\
  & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
  & \Prop^\sim\ p\ P\ Q = ...
\end{align*}

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.

\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t :\equiv
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

Note that contractibility of singletons holds in SeTT by UIP and the computation
rules of equality on $\Sigma$ types. Hence, the J eliminator can also be
derived.

@@

In its current form, SeTT is already rich and interesting in its own right.
However, as pointed out in the introduction, it is also seriously limited by the
lack of a universe internalizing the notion of types beyond propositions, that
is, a universe of setoids. Our goal is to extend SeTT with such universe; since
SeTT is a direct syntactic reflection of the setoid model, this essentially
amounts to showing that a universe of setoids with the necessary structure and
equations can be constructed within the setoid model. This opens several
questions and possible design choices.

A first fundamental consideration has to with the very definition of the setoid
universe: as any type in the setoid model, this universe must be a setoid and
thus come equipped with an equivalence relation. However, unlike the already
existing universe of propositions, a universe of setoids cannot be univalent,
since this would force it to be a groupoid. Instead, we can define the relation
so that it reflects a simpler syntactic equality of codes.

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section TODO and shown originally in [cite LICS
  99] is that it can be carried out in a very small type theoretic metatheory,
thus providing a way to reduce extensionality to a small intensional core.
%
In extending the setoid model with a universe, we would like this property to
still hold, hence it is of primary concern to reduce to a minimum the extensions
required in the metatheory, so to keep it as small as possible.

A known and established method for constructing universes in Type Theory uses
induction-recursion (IR), a definition schema developed by Dybjer
[cite]. Inductive-recursive definitions can be found throughout the literature
on type theory, from the already mentioned type theoretic universes, including
the original formulation a la Tarski by Martin-L\"of [cite], to metamathematical
tools like computability predicates [cite].

Although universe constructions in Type Theory---including our own setoid
universe---are naturally presented as inductive-recursive definitions, they
may not necessarily require a metatheory that supports induction-recursion. In
fact, it is possible to reduce some instances of induction-recursion to plain
induction (more specifically, inductive families), including some universe
definitions. We recall this reduction in Section TODO.

%% However, we would prefer not to be forced to assume inductive-recursive types,
%% or IR in general, in the basic type theory which is the target of the setoid
%% translation. This is because the point of the translation is to explain a
%% complex system in terms of a very simple, well understood core, and we think
%% introducing IR would go against this goal.

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support elimination, thus exposing the inductive structure of
the codes. Universe elimination is known to be inconsistent with univalence. As
it turns our, univalence is not an issue in our case; nevertheless it is not
immediately clear if the elimination principle can be justified by the
semantics.

Another choice is whether to include codes for other universes, besides the
usual type formers. One could have $\Prop$ be a code in the setoid universe,
rather than a type. Moreover, the universes could form a hierarchy, where the
universe at level $i$ is an element of the universe of level $i+1$. Another
option is whether to make the hierarchy cumulative.

%% In the next section, we recall the inductive-recursive construction of universes
%% in type theory, as well as lay out the process by which such construction can be
%% turned to one that only relies on inductive families. In Section TODO, we give a
%% first inductive-recursive definition of the universe of setoids. After that we
%% will see if and how such IR definition can be turned into one that does not rely
%% on IR.

@@

An inductive-recursive universe is given by a type of codes $\U$, and a family
$\El : \U \to \mType$ that assigns, to each code corresponding to some type, the
meta-theoretic type of its elements. The resulting definition is
inductive-recursive because the inductive type of codes is defined
simultaneously with the recursive function $\El$.

As an example, consider the following definition of a small universe with
booleans and $\Pi$ types.

\begin{align*}
  & \textsf{data}\ \U\ : \mType_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

Note that $\U$ is defined to be in $\mType_1$ rather than $\mType$. This is
because... \todo{TODO add explanation}. This fact also implies that the
meta-theory must contain at least one universe.

Induction-recursion is arguably the nicest and more natural way to define
internal universes in Type Theory, however it is not always required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of sets and sets over $I$,
i.e.\ $[I, \mType_i] \simeq \mType_i/I$ for $I : \mType_i$.

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$. In other words, $\ad{in-U}$ is a predicate that holds for
any type that would have been obtained via $\El$ in the usual
inductive-recursive definition. As $\El$ is indexed on the type of codes, the
definition of $\ad{in-U}$ quite expectedly reflecs the inductive structure of
codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\Set$, since the definition of $\U$ quantifies over all possible sets.

@@

%% \todo{QIIT II LICS paper. this part is already in the HIIT paper}
